A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-06-27 tex/sec4-pre-synthesis.tex
[ tex ] Update the discussion about annotatability

\citet{Dunfield2021} formulated completeness differently from ours (\cref{thm:completeness}) and proposed \emph{annotatability} as a more suitable name.
Here we reformulate annotatability in our theory and discuss its relationship with our completeness and generalised mode preprocessing.

@@

Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term is typable in $(\Sigma, \erase\Omega)$, it may not be typable in $(\Sigma, \Omega)$ directly, but will be if some annotations are added.
Indeed, \cref{thm:annotatability} can be easily proved in our theory from generalised mode preprocessing.

\begin{proof}[Proof of \cref{thm:annotatability}]
By \cref{thm:generalised-mode-preprocessing}, there is a generalised mode derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some mode~$\dir{d}$ and booleans $g$~and~$s$.
Perform induction on this generalised mode derivation to construct a bidirectional typing derivation in the same mode (and also the term~$t'$ and a proof of $t' \sqsupseteq t$, which are straightforward and omitted here).
\begin{itemize}
\item $\SynRule{Var}$, $\SynRule{Anno}$, and $\Rule{Op}$:
The outermost rule of the given typing derivation must be the corresponding one; apply the corresponding bidirectional typing rule to the induction hypotheses.
\item $\ChkRule{Sub}$:
Apply $\ChkRule{Sub}$ to the induction hypothesis.
\item $\SynRule{Missing}$:
This is the interesting case where we map a $\SynRule{Missing}$ rule to an $\SynRule{Anno}$ rule and add to the term a type annotation, which comes from the given typing derivation.
\vspace{-\topsep-\baselineskip}
\end{itemize}
\end{proof}

On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), annotatability cannot replace our completeness since it relates typing derivations for possibly differently annotated terms, but a bidirectional type synthesiser is concerned about properties of \emph{one} given term throughout the process.
Annotatability also does not help the user to deal with missing annotations like generalised mode preprocessing does: even though annotatability seems to construct a term with missing annotations filled in, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
It seems that as we focus on properties that serve practical purposes, it becomes unclear how annotatability comes into the picture.
In particular, annotatability mixes up the aspects of annotating and typing, and therefore does not fit into our theory, where the two aspects are separated for better usability.

>>>>

\citet{Dunfield2021} formulated completeness differently from ours (\cref{thm:completeness}) and proposed \emph{annotatability} as a more suitable name.
Here we formulate annotatability in our theory and discuss its relationship with our completeness and generalised mode preprocessing.

@@

Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term is typable in $(\Sigma, \erase\Omega)$, it may not be typable in $(\Sigma, \Omega)$ directly due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be easily proved from generalised mode preprocessing.

\begin{proof}[Proof of \cref{thm:annotatability}]
By \cref{thm:generalised-mode-preprocessing}, there is a generalised mode derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some mode~$\dir{d}$ and booleans $g$~and~$s$.
Perform induction on this generalised mode derivation to construct a bidirectional typing derivation in the same mode (and also the term~$t'$ and a proof of $t' \sqsupseteq t$, which are determined by the derivations and omitted here).
\begin{itemize}
\item $\SynRule{Var}$, $\SynRule{Anno}$, and $\Rule{Op}$:
The outermost rule of the given typing derivation must be the corresponding one; apply the corresponding bidirectional typing rule to the induction hypotheses.
\item $\ChkRule{Sub}$:
Apply $\ChkRule{Sub}$ to the induction hypothesis.
\item $\SynRule{Missing}$:
This is the interesting case where we map a $\SynRule{Missing}$ rule to an $\SynRule{Anno}$ rule and add to the term a type annotation, which comes from the given typing derivation.
\vspace{-\topsep-\baselineskip}
\end{itemize}
\end{proof}

On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), our completeness is much simpler to use than annotatability, which requires the bidirectional type synthesiser to produce more complex evidence when the synthesis fails.
Annotatability also does not help the user to deal with missing annotations like (generalised) mode preprocessing does: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
Therefore we believe that the notion of annotatability is too complex for what it achieves, and our theory offers simpler and more useful alternatives.
